\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in \proglang{R}

### 2014-11-26 papers/jss/article.Rnw
Cut section 6 in half by removing the section explaining the caveats about formulas and types not supported by serialized_pb, and just explain now that we serialize everything, but in one sentence explain the caveat that we fall back to base::serialize for R-specific types like language,function,and environment.  This basically removes the need for 6.1 at all, so remove that section, and move a tiny bit of the top text about which datasets we are using into the top of section 6.2 which explains the compresison performance.
Next step: Replace table with a plot as hadley and one of the referees

The \code{rexp.proto} schema supports all main \proglang{R} storage types holding \emph{data}.
These include \code{NULL}, \code{list} and vectors of type \code{logical}, 
\code{character}, \code{double}, \code{integer}, and \code{complex}. In addition,
every type can contain a named set of attributes, as is the case in \proglang{R}. The \code{rexp.proto}
schema does not support some of the special \proglang{R} specific storage types, such as \code{function},
\code{language} or \code{environment}. Such objects have no native equivalent 
type in Protocol Buffers, and have little meaning outside the context of \proglang{R}.
When serializing \proglang{R} objects using \code{serialize\_pb}, values or attributes of
unsupported types are skipped with a warning. If the user really wishes to serialize these 
objects, they need to be converted into a supported type. For example, the  can use 
\code{deparse} to convert functions or language objects into strings, or \code{as.list}
for environments.

\subsection[Evaluation: Converting R data sets]{Evaluation: Converting \proglang{R} data sets}

To illustrate how this method works, we attempt to convert all of the built-in 
data sets from \proglang{R} into this serialized Protocol Buffer representation.

<<echo=TRUE>>=
datasets <- as.data.frame(data(package="datasets")$results)
datasets$name <- sub("\\s+.*$", "", datasets$Item)
n <- nrow(datasets)
@

There are \Sexpr{n} standard data sets included in the \pkg{datasets}
package included with \proglang{R}. These data sets include data frames, matrices, time series, tables lists,
and some more exotic data classes. The \code{can\_serialize\_pb} method is 
used to determine which of those can fully be converted to the \code{rexp.proto}
Protocol Buffer representation. This method simply checks if any of the values or
attributes in an object is of an unsupported type:

<<echo=TRUE>>=
m <- sum(sapply(datasets$name, function(x) can_serialize_pb(get(x))))
@

\Sexpr{m} data sets can be converted to Protocol Buffers
without loss of information (\Sexpr{format(100*m/n,digits=1)}\%). Upon closer
inspection, all other data sets are objects of class \code{nfnGroupedData}.
This class represents a special type of data frame that has some additional 
attributes (such as a \emph{formula} object) used by the \pkg{nlme} package \citep{nlme}.
Because formulas are \proglang{R} \emph{language} objects, they have little meaning to
other systems, and are not supported by the \code{rexp.proto} descriptor.
When \code{serialize\_pb} is used on objects of this class, it will serialize
the data frame and all attributes, except for the formula.

<<>>=
attr(CO2, "formula")
msg <- serialize_pb(CO2, NULL)
object <- unserialize_pb(msg)
identical(CO2, object)
identical(class(CO2), class(object))
identical(dim(CO2), dim(object))
attr(object, "formula")
@

\subsection{Compression performance}
\label{sec:compression}

This section compares how many bytes are used to store data sets
using four different methods:

@@

Table~\ref{tab:compression} shows the sizes of 50 sample \proglang{R} data sets as
returned by object.size() compared to the serialized sizes.
%The summary compression sizes are listed below, and a full table for a
%sample of 50 data sets is included on the next page.  
Note that Protocol Buffer serialization results in slightly
smaller byte streams compared to native \proglang{R} serialization in most cases,
but this difference disappears if the results are compressed with gzip.
%Sizes are comparable but Protocol Buffers provide simple getters and setters
%in multiple languages instead of requiring other programs to parse the \proglang{R}
%serialization format. % \citep{serialization}.
One takeaway from this table is that the universal \proglang{R} object schema
included in \pkg{RProtoBuf} does not in general provide
any significant saving in file size compared to the normal serialization
mechanism in \proglang{R}.
% redundant: which is seen as equally compact.
The benefits of \pkg{RProtoBuf} accrue more naturally in applications where
multiple programming languages are involved, or when a more concise
application-specific schema has been defined.  The example in the next
section satisfies both of these conditions.

@@

\begin{verbatim}
package rexp;

message REXP {
  enum RClass {
    STRING = 0;
    RAW = 1;
    REAL = 2;
    COMPLEX = 3;
    INTEGER = 4;
    LIST = 5;
    LOGICAL = 6;
    NULLTYPE = 7;
  }
  enum RBOOLEAN {
    F=0;
    T=1;
    NA=2;
  }

  required RClass rclass = 1 ; 
  repeated double realValue = 2 [packed=true];
  repeated sint32 intValue = 3 [packed=true];
  repeated RBOOLEAN booleanValue = 4;
  repeated STRING stringValue = 5;
  optional bytes rawValue = 6;
  repeated CMPLX complexValue = 7;
  repeated REXP rexpValue = 8;
  repeated string attrName = 11;
  repeated REXP attrValue = 12;
}
message STRING {
  optional string strval = 1;
  optional bool isNA = 2 [default=false];
}
message CMPLX {
  optional double real = 1 [default=0];
  required double imag = 2;
}
\end{verbatim}
% \end{appendices}
\newpage
\bibliography{article}

>>>>

The \code{rexp.proto} schema natively supports all main \proglang{R}
storage types holding \emph{data}.  These include \code{NULL},
\code{list} and vectors of type \code{logical}, \code{character},
\code{double}, \code{integer}, and \code{complex}. In addition, every
type can contain a named set of attributes, as is the case in
\proglang{R}. The storage types \code{function}, \code{language}, and
\code{environment} are specific to \proglang{R} and have no equivalent
native type in Protocol Buffers.  These three types are supported by
first serializing with \code{base::serialize} in \proglang{R} and
then stored in a raw bytes field.

\subsection[Evaluation: Serializing R data sets]{Evaluation: Serializing \proglang{R} data sets}
\label{sec:compression}

<<echo=FALSE>>=
datasets <- as.data.frame(data(package="datasets")$results)
datasets$name <- sub("\\s+.*$", "", datasets$Item)
n <- nrow(datasets)
@

This section evaluates the effectiveness of serializing arbitrary
\proglang{R} data structures into Protocol Buffers.  We use the
\Sexpr{n} standard data sets included in the \pkg{datasets} package
included with \proglang{R} as our evaluation data. These data sets
include data frames, matrices, time series, tables, lists, and some
more exotic data classes.  For each data set, we compare how many
bytes are used to store the data set using four different methods:

@@

Table~\ref{tab:compression} shows the sizes of 50 sample \proglang{R} data sets as
returned by \code{object.size()} compared to the serialized sizes.
%The summary compression sizes are listed below, and a full table for a
%sample of 50 data sets is included on the next page.
Note that Protocol Buffer serialization results in slightly
smaller byte streams compared to native \proglang{R} serialization in most cases,
but this difference disappears if the results are compressed with gzip.
%Sizes are comparable but Protocol Buffers provide simple getters and setters
%in multiple languages instead of requiring other programs to parse the \proglang{R}
%serialization format. % \citep{serialization}.
One takeaway from this table is that the universal \proglang{R} object schema
included in \pkg{RProtoBuf} does not in general provide
any significant saving in file size compared to the normal serialization
mechanism in \proglang{R}.
% redundant: which is seen as equally compact.
The benefits of \pkg{RProtoBuf} accrue more naturally in applications where
multiple programming languages are involved, or when a more concise
application-specific schema has been defined.  The example in the next
section satisfies both of these conditions.

@@

\begin{verbatim}
package rexp;
message REXP {
  enum RClass {
    STRING = 0;
    RAW = 1;
    REAL = 2;
    COMPLEX = 3;
    INTEGER = 4;
    LIST = 5;
    LOGICAL = 6;
    NULLTYPE = 7;
    LANGUAGE = 8;
    ENVIRONMENT = 9;
    FUNCTION = 10;
  }
  enum RBOOLEAN {
    F=0;
    T=1;
    NA=2;
  }
  required RClass rclass = 1;
  repeated double realValue = 2 [packed=true];
  repeated sint32 intValue = 3 [packed=true];
  repeated RBOOLEAN booleanValue = 4;
  repeated STRING stringValue = 5;
  optional bytes rawValue = 6;
  repeated CMPLX complexValue = 7;
  repeated REXP rexpValue = 8;
  repeated string attrName = 11;
  repeated REXP attrValue = 12;
  optional bytes languageValue = 13;
  optional bytes environmentValue = 14;
  optional bytes functionValue = 14;
}
message STRING {
  optional string strval = 1;
  optional bool isNA = 2 [default=false];
}
message CMPLX {
  optional double real = 1 [default=0];
  required double imag = 2;
}
\end{verbatim}
% \end{appendices}
\newpage
\bibliography{article}
